Structure Learning for $\VAR(p)$ Models

### 2019-06-25 paper/paper.tex
Cleanup on statistical testing portion of paper.

\begin{align}
  H_0:&\ \widehat{x}_i^{(p_i)}(t) = \sum_{\tau = 1}^{p_i} b_{ii}(\tau)x_i(t - \tau),\\
  H_1:&\ \widehat{x}_i^{(p_{ij})}(t) = \sum_{\tau = 1}^{p_{ij}} b_{ii}(\tau)x_i(t - \tau) + \sum_{\tau = 1}^{p_{ij}}b_{ij}(\tau)x_j(t - \tau).
\end{align}

Under the null hypothesis $H_0$ (i.e. that $x_j \npwgc x_i$) we can
form the asymptotically $\chi^2(p_{ij})$ distributed test statistic
\todo{Is this actually $\chi^2$ when $p_{ij}$ is chosen from the data?  This is adaptive!}

\begin{equation}
  \label{eqn:gc_statistics}
  F_{ij}(p_i, p_{ij}) = \frac{T}{p_{ij}}\Big(\frac{\xi_i(p_i)}{\xi_{ij}(p_{ij})} - 1\Big),
\end{equation}

where $\xi_i(p_i)$ is the sample mean square of the residuals\footnote{This
  quantity is often denoted $\widehat{\sigma}$, but we maintain
  notation from Definition \ref{def:granger_causality}.}
$x_i(t) - \widehat{x}^{(p_i)}_i(t)$

\begin{equation*}
  \xi_i(p_i) = \frac{1}{T - p_i}\sum_{t = p_i + 1}^T (x_i(t) - \widehat{x}_i^{(p_i)}(t))^2,
\end{equation*}

and similarly for $\xi_{ij}(p_{ij})$.  Under moment conditions $x(t)$
(\hl{TODO: What assumptions? -- this must also depend on choice of
  $p_i, p_j$}) we have the convergence in probability:

\begin{equation}
  F_{ij}(p_i, p_{ij}) \overset{\P}{\rightarrow}
  \left\{
    \begin{array}{ll}
      0;\ x_j \npwgc x_i\\
      \infty;\ x_j \pwgc x_i
    \end{array}
  \right. \text{ as } T \rightarrow \infty.  % the '.' after \right is necessary.
\end{equation}

Therefore, we can make the statement that Algorithm
\ref{alg:finite_pwgc} asymptotically implements Algorithm
\ref{alg:pwgr}.  \hl{TODO: Modify Algorithm 2 to make this true.}
\hl{TODO: Enclose this into a theorem}

\subsection{Model Order Selection}
\label{sec:model_order_selection}
There are a variety of methods to choose the filter orders $p_i$ and
$p_j$ (see e.g. \cite{lutkepohl2005new}), but we will focus in
particular on the Bayesian Information Criteria (BIC).  The BIC is
substantially more conservative than the popular alternative Akaiake
Information Criteria (the BIC is also asymptotically consistent), and
since we are searching for \textit{sparse graphs}, we therefore prefer
the BIC, where we seek to \textit{minimize} over $p$:

@@

We compare our results against the LASSO as a baseline, estimating the
filter $\B(z)$ that minimizes $\xi_{\text{LASSO}}$ in the following:

\begin{equation}
  \begin{aligned}
  \xi_{\text{LASSO}}^{(\lambda)} &= \underset{B}{\text{min}}\ \frac{1}{T}\sum_{t = p + 1}^T||x(t) - \sum_{\tau = 1}^p B(\tau) x(t - \tau)||_2^2 + \lambda \sum_{\tau = 1}^p ||\matvec B(\tau)||_1\\
  \xi_{\text{LASSO}} &= \underset{\lambda \ge 0}{\text{min}}\ \xi_{\text{LASSO}}^{(\lambda)} + \mathsf{BIC}(B_{\text{LASSO}}^{(\lambda)})\\
  \end{aligned}
\end{equation}

where $||\matvec B||_1$ emphasizes the application of a vector norm, and
$\lambda$ is the regularization parameter chosen via the BIC.

@@

In reference to figure \ref{fig:simulation_results_comparison1} it
should not be overly surprising that our PWGC algorithm performs
better than the LASSO for the case of a strongly causal graph, since
in this case the assumptions which guarantee the correctness of
Algorithm \ref{alg:finite_pwgc} hold.  However, the performance is
still markedly superior in the case of a more general DAG.

In regards scalability, we have observed that performing the $O(n^2)$
pairwise Granger-causality calculations consumes the vast majority
($> 90\%$) of the computation time.  Since this step is trivially
parallelizable, our algorithm also scales well with multiple cores or
multiple machines.  Figure \ref{fig:simulation_results_scaling} is a
demonstrating of this scalability, where we are able to estimate
graphs having over $1500$ nodes (over $2.25 \times 10 ^6$ possible
edges) using only $T = 500$ data points, granted, an SCG on this many
nodes is extremely sparse.  Application to denser graphs is explored
in Figure \ref{fig:simulation_results_dense}, where $n$ and $T$ are
held fixed while increasing the edge probability $q$, thereby creating
denser graphs.  For larger values of $q$ in this plot there is a
comparable number of edges as there are data samples $T$.

>>>>

\begin{align}
  H_0:&\ \widehat{x}_i^{(p)}(t) = \sum_{\tau = 1}^{p} b_{ii}(\tau)x_i(t - \tau),\\
  H_1:&\ \widehat{x}_i^{(p)}(t) = \sum_{\tau = 1}^{p} b_{ii}(\tau)x_i(t - \tau) + \sum_{\tau = 1}^pb_{ij}(\tau)x_j(t - \tau).
\end{align}

We formulate the statistic

\begin{equation}
  \label{eqn:gc_statistics}
  F_{ij}(p) = \frac{T}{p}\Big(\frac{\xi_i(p)}{\xi_{ij}(p)} - 1\Big),
\end{equation}

where $\xi_i(p)$ is the sample mean square of the
residuals\footnote{This quantity is often denoted $\widehat{\sigma}$,
  but we maintain notation from Definition
  \ref{def:granger_causality}.}  $x_i(t) - \widehat{x}^{(p)}_i(t)$,

\begin{equation*}
  \xi_i(p) = \frac{1}{T - p}\sum_{t = p + 1}^T (x_i(t) - \widehat{x}_i^{(p)}(t))^2,
\end{equation*}

and similarly for $\xi_{ij}(p)$.  We test $F_{ij}(p)$ against a
$\chi^2(p)$ distribution.

If disturbances are Gaussian, we will have the convergence

\begin{equation}
  F_{ij}(p) \overset{\P}{\rightarrow}
  \left\{
    \begin{array}{ll}
      0;\ x_j \npwgc x_i\\
      \infty;\ x_j \pwgc x_i
    \end{array}
  \right. \text{ as } T \rightarrow \infty.  % the '.' after \right is necessary.
\end{equation}

In our finite sample implementation (see Algorithm
\ref{alg:finite_pwgc}) we add edges to $\widehat{\gcg}$ in order of
the decreasing magnitude of $F_{ij}$ instead of proceeding backwards
through $P_{k - r}$ in Algorithm \ref{alg:pwgr}.  This makes greater
use of the information provided by the test statistic $F_{ij}$,
moreover, if $x_i \gc x_j$ and $x_j \gc x_k$, it is expected that
$F_{kj} > F_{ji}$, thereby providing the same effect as proceeding
backwards through $P_{k - r}$.
\subsection{Model Order Selection}
\label{sec:model_order_selection}
There are a variety of methods to choose the filter orders $p_i$ and
$p_j$ (see e.g. \cite{lutkepohl2005new}), but we will focus in
particular on the Bayesian Information Criteria (BIC).  The BIC is
substantially more conservative than the popular alternative Akaiake
Information Criteria (the BIC is also asymptotically consistent), and
since we are searching for \textit{sparse graphs}, we therefore prefer
the BIC, where we seek to \textit{minimize} over $p$:

@@

We compare our results against the LASSO.  Motivated by scaling, we
split the squared error term into separate terms, one for each group of
incident edges on a node, and estimate the collection of $n$ incident
filters $\big\{\B_{ij}(z)\big\}_{j = 1}^n$ that minimizes
$\xi_i^{\text{LASSO}}$ in the following:

\begin{equation}
  \begin{aligned}
  \xi_i^{\text{LASSO}}(\lambda) &= \underset{B}{\text{min}}\ \frac{1}{T}\sum_{t = p + 1}^T\big(x_i(t) - \sum_{\tau = 1}^p\sum_{j = 1}^n B_{i, j}(\tau) x(t - \tau)\big)^2 + \lambda \sum_{\tau = 1}^p \sum_{j = 1}^n |B_{ij}(\tau)|\\
  \xi_i^{\text{LASSO}} &= \underset{\lambda \ge 0}{\text{min}}\ \xi_i^{\text{LASSO}}(\lambda) + \mathsf{BIC}\big(B_i^{\text{LASSO}}(\lambda)\big)\\
  \end{aligned}
\end{equation}

where we are choosing $\lambda$, the regularization parameter, via the BIC.

@@

In reference to figure \ref{fig:simulation_results_comparison1} it
should not be overly surprising that our PWGC algorithm performs
better than the LASSO for the case of a strongly causal graph, since
in this case the assumptions which guarantee the correctness of
Algorithm \ref{alg:finite_pwgc} hold.  However, the performance is
still markedly superior in the case of a more general DAG.  We would
conjecture that a DAG having a similar degree of sparsity as an SCG is
likely to be ``close'' to an SCG.  Figure
\ref{fig:simulation_results_dense} illustrates the severe (expected)
degredation in performance as the number of DAG edges increases while
the number of data samples $T$ remains fixed.  For larger values $q$
in this plot, the number of edges in the graph is comparable to the
number of data samples.

In regards scalability, we have observed that performing the $O(n^2)$
pairwise Granger-causality calculations consumes the vast majority
($> 90\%$) of the computation time.  Since this step is trivially
parallelizable, our algorithm also scales well with multiple cores or
multiple machines.  Figure \ref{fig:simulation_results_scaling} is a
demonstrating of this scalability, where we are able to estimate
graphs having over $1500$ nodes (over $2.25 \times 10 ^6$ possible
edges) using only $T = 500$ data points, granted, an SCG on this many
nodes is extremely sparse.  However, it is certainly possible to
improve the scalability of the LASSO as applied to time series data by
taking advantage of the block-Toeplitz covariance structure.
